Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-09 paper/paper.tex

\section{Introduction}
\label{sec:intro}
Structured recursion schemes~\cite{HinzeW16,HinzeWG15} are powerful
abstractions that capture common patterns of recursion. The main benefit of
structuring computation using recursion schemes, is that they enjoy
well-estabished \emph{algebraic properties} that can serve as a foundation for
reasoning about program equivalences, transformations, and
optimisations (e.g.\ \emph{fusion} laws, or semi-automatic
parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}).
These algebraic properties led to their use in the context of
\emph{program calculation}, where programmers would describe their code using
simple, inefficient specifications in an \emph{algebra of
programming}~\cite{BirddeMoor96:Algebra}, and then use the algebraic laws of
this algebra of programming to \emph{calculate} an efficient version of the same
algorithm. Suppose, for example, that we want to write a program that sorts a list of
integers, and then multiplies by 2 all its elements. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}
Instead of writing this function directly, we may observe that we can derive it
by \emph{fusing} a regular \emph{quickort} OCaml implementation, with
\ocaml{map (fun x -> x * 2)}. This is the main idea behind program calculation:
start
from a simple specification, e.g.\
$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
equivalences and algebraic laws to rewrite it to an optimised version  (e.g.\
the fused OCaml implementation above). We will revisit a similar example
in Section~\ref{sec:sorting}.

@@

\section{Recursion Schemes}\label{sec:rec-schemes}
The structure of data is very similar to the structure of an
algorithm which processes that data. This relationship manifests in the form of
structured recursion schemes which are widely used in functional languages such
Haskell. Canonical examples are folds (catamorphisms), which consume data, and
unfolds (anamorphisms), which produce it. While some implementations of
recursion schemes like $\operator{foldr}$ in Haskell are specific to a particular
data structure, in this case $\operator{List}$s, we can generalise further these
ideas to account for generic (co)inductive data types and generic algorithms
operating on them. This is done by looking at recursion schemes from the point
of view of category theory.

@@

For example, the category of sets, denoted by $\Set$, is the category where
objects are sets and arrows are functions between sets. The initial object $0$
in $\Set$ is the empty set $\emptyset$ and the terminal object $1$ is any
singleton set. In this section, we are going to assume types are sets, hence
this category can also be viewed as the category of types and programs between
them.

@@

For example, in $\Set$, we can define the set of lists
$\List{A} \iso 1 + A \times \List{A}$ as the set inductively generated by the
constructors $\nil : 1 \to \List{A}$ and
$\cons : A \times \List{A} \to \List{A}$.
The terminal object in $\Set$ is the \emph{unit} type, that only has one
inhabitant, $* : 1$, analogous to the unit type in Haskell: \haskell{() :: ()}.
The $\List{-}$ type is a functor. Its action on objects is to take any set
$A$ to $\List{A}$, and its functorial
action $\List{f} : \List{A} \to \List{B}$ is given by $\List{f}(*) = *$ and
$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
$\List{f}$ is well-defined as it recursively calls on a smaller argument.
However, there is a more general way of proving these type of definitions are
well-defined. Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
the greatest set generated by the constructor
$\cons : A \times \Str{A} \to \Str{A}$.

@@

@@

>>>>

\section{Introduction}
\label{sec:intro}
Structured recursion schemes~\cite{HinzeW16,HinzeWG15} are powerful
abstractions that capture common patterns of recursion. The main benefit of
structuring computation using recursion schemes, is that they enjoy
well-estabished \emph{algebraic properties} that can serve as a foundation for
reasoning about program equivalences, transformations, and
optimisations (e.g.\ \emph{fusion} laws, or semi-automatic
parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}).
These algebraic properties led to their use in the context of
\emph{program calculation}, where programmers would describe their code using
simple, inefficient specifications in an \emph{algebra of
programming}~\cite{BirddeMoor96:Algebra}, and then use the algebraic laws of
this algebra of programming to \emph{calculate} an efficient version of the same
algorithm. Suppose, for example, that we want to write a program that sorts a list of
integers, and then multiplies by 2 all its elements. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}
Instead of writing this function directly, we may observe that we can derive it
by \emph{fusing} a regular \emph{quicksort} OCaml implementation, with
\ocaml{map (fun x -> x * 2)}. This is the main idea behind program calculation:
start
from a simple specification, e.g.\
$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
equivalences and algebraic laws to rewrite it to an optimised version  (e.g.\
the fused OCaml implementation above). We will revisit a similar example
in Section~\ref{sec:sorting}.

@@

\section{Recursion Schemes}\label{sec:rec-schemes}
The structure of data is very similar to the structure of an
algorithm which processes that data. This relationship manifests in the form of
structured recursion schemes which are widely used in functional languages such
Haskell. Canonical examples are folds (catamorphisms), which consume data, and
unfolds (anamorphisms), which produce it. While some implementations of
recursion schemes like $\operator{foldr}$ in Haskell are specific to a particular
data structure, in this case $\operator{List}$s, we can generalise further these
ideas to account for generic (co)inductive data types and generic algorithms
operating on them.

Furthermore folds and unfolds can be shown to capture a wide
range of recursion schemes such as primitive recursion, mutual recursion,
dynamic programming algorithms, polymorphic recursion, recursion with
accumulators and so on. However, for divide-and-conquer algorithms folds need to
be generalised to hylomorphisms which provide the ultimate basic building block
for any other recursion scheme. This is done by looking at recursion schemes
from the point of view of category theory.

@@

For example, the category of sets, denoted by $\Set$, is the category where
objects are sets and arrows are functions between sets. The initial object $0$
in $\Set$ is the empty set $\emptyset$ and the terminal object $1$ is any
singleton set. The reader who is not accustomed with category theory can assume
types are sets, giving the intuition that the category $\Set$ can also be viewed
as the category of (simple) types and programs between them.

@@

In $\Set$, we can define the set of lists
\[
  \List{A} \iso 1 + A \times \List{A}
\]
as the set inductively generated by the
constructors $\nil : 1 \to \List{A}$ and
$\cons : A \times \List{A} \to \List{A}$.
The terminal object in $\Set$ is the \emph{unit} type, that only has one
inhabitant, $* : 1$, analogous to the unit type in Haskell: \haskell{() :: ()}.
The $\List{-}$ type is a functor. Its action on objects is to take any set
$A$ to $\List{A}$, and its functorial
action $\List{f} : \List{A} \to \List{B}$ is given by $\List{f}(*) = *$ and
$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
$\List{f}$ is well-defined as it recursively calls on a smaller argument.
However, there is a more general way of proving these type of definitions are
well-defined. Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
the greatest set generated by the constructor
$\cons : A \times \Str{A} \to \Str{A}$.

@@

In order to find a solution we restrict ourselves to the so-called
\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.
A $F$-coalgebra $(X, c)$ is recursive if

An example of a
recursive coalgebra is the \coq{partition} function in quicksort which
destructures a list into a pivot and two sublist and as long as the sublists are
smaller the partitioning function still yields a unique solution to the
recursion scheme.

\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
\end{coqcode}
Using the hylo fusion law, we can prove the well-known \emph{deforestation
optimisation}. This is when two consecutive recursive computations, one that
builds a data structure, and another one that consumes it, can be fused together
into a single recursive definition. This, in turn, allows us to prove that a
recursive hylomorphism is the composition of a catamorphism and a recursive
anamorphism.
\begin{coqcode}
Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
  (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
  : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.
\end{coqcode}

\subsection{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way.  Intuitively, a recursive coalgebra
can be applied only finitely many times, therefore when this is applied to an
anamorphism the only possible traces it can produce from the seed function are
the finite ones. We denote this special kind of recursion scheme \emph{recursive
anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
from the universality property of the hylomorphism and the fact that recursive
anamorphisms satisfy the same equation.

\subsection{Haskell Implementation of Recursion Schemes}
\begin{minted}{haskell}
newtype Fix f = In {inOp :: f (Fix f )}
\end{minted}

\begin{minted}{haskell}
cata :: Functor f => (f a -> a) -> Fix f -> a
cata alg = alg . fmap (cata alg) . In
\end{minted}
\begin{minted}{haskell}
foldr :: (a -> b -> b) -> b -> [ a ] -> b
foldr f e [ ]    = e
foldr f e (x:xs) = f x (foldr f e xs)
\end{minted}
which is very useful in list processing. As a fold, The frequently used function map is also a fold:
\begin{minted}{haskell}
map :: (a -> b) -> [a] -> [b]
map f = foldr (\x xs -> (f x):xs) []
\end{minted}

\begin{minted}{haskell}
newtype CoFix f = OutOp {out :: f (Fix f )}
\end{minted}

\begin{minted}{haskell}
ana :: Functor b => (a -> b a) -> a -> CoFix b
ana coalg = OutOp . fmap (ana coalg) . coalg
\end{minted}

 and can be
thought of as the $\operator{fmap}$ program in functional programming.
\begin{minted}{haskell}
class Functor f where
  fmap :: (a -> b) -> f a -> f b
\end{minted}

@@

\subsubsection{On the subtype of finite elements}
In this development we have defined recursive anamorphisms on inductive data
types. We might have as well defined them on the subtype of finite elements of
coinductive data types using a predicate which states when an element of a
coinductive data type is finite:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
is isomorphic its corresponding the inductive data type \coq{LFix F}. This is
easy to see. We first define a catamorphism \coq{ccata_f_} from the
subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
any \coq{F}-algebra.
\begin{coqcode}
Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
  : forall x : GFix F, FinF x -> A := fix f x H :=
    let hx := g_out x in
      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
\end{coqcode}
We now prove this is isomorphic to the least fixed-point of the functor
\coq{F}.
We take the catamorphism from the finite elements of \coq{GFix F} to the
inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its
inverse is the catamorphism on the restriction of \coq{g_in} to the finite
elements of \coq{GFix}, which we denote by \coq{lg_in}.  The following lemmas
prove the isomorphism:
\begin{coqcode}
Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
Lemma ccata_cata `{setoid A} : ccata l_in \o cata lg_in =e id.
\end{coqcode}
The finite subtype of \coq{GFix F} allows us to compose
catamorphisms and anamorphisms, by using the above isomorphism. In our work,
however, we use \emph{recursive} anamorphisms, defined as \coq{hylo l_in c} for
a recursive coalgebra \coq{c}, which compose easily with
catamorphisms.
